DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-03-04 manuscript/manuscript.Rmd
figure 1 vertical mostly final

Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals [@Kumar2017].
Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education [@webb2005phylomatic; @Stoltzfus2013].
<!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [@].
The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
Here we present the full implementation of DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/.
The current implementation of the `datelife` R package is available on CRAN [@] and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize and compare source chronograms
<!--, Functions to visualize and graphically compare source and summary chronograms are available at `datelifeplot` not on CRAN yet.-->

@@

1. It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths.
1. Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, `datelife` implements the Taxonomic Name Resolution Service [TNRS; @Boyle2013] provided by OpenTree [@opentreeAPIs]. TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
1. The current version of `datelife` only accepts scientific taxonomic names as input. Names can belong to any taxonomic group or binomial specific. If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
1. The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided. The \texttt{datelifeQuery} object is used next to search the chronogram database.
1. Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips. Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
1. At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
1. To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree. Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages. Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
1. Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.

@@

At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem]. A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN [@]. `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.

>>>>

Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals [@Kumar2017].
Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education [@webb2005phylomatic; @Stoltzfus2013].
<!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user.
The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
Here we present the full implementation of DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/.
The latest implementation of the `datelife` R package is available on CRAN (v`r packageVersion("datelife")`; @datelifeR) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize chronograms and compare them.
<!--, Functions to visualize and graphically compare source and summary chronograms are available at `datelifeplot` not on CRAN yet.-->

@@

1. It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree. If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004]. The input tree is not required to have branch lengths.
1. Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, `datelife` implements the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs]. TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
1. The current version of `datelife` only accepts scientific taxonomic names as input. Names can belong to any taxonomic group or binomial specific. If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag. If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input. If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
1. The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided. The \texttt{datelifeQuery} object is used next to search the chronogram database.
1. Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips. Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
1. At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
1. To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree. Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages. Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
1. Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.

@@

At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem]. A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN. `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.

### 2022-03-04 manuscript/manuscript.tex
figure 1 vertical mostly final

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3670}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals (Kumar, Stecher, Suleski, \& Hedges, 2017).
Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O. Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education (Stoltzfus et al., 2013; Campbell O. Webb \& Donoghue, 2005).
We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}@{]}.
The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
Here we present the full implementation of DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/.
The current implementation of the \texttt{datelife} R package is available on CRAN {[}@{]} and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize and compare source chronograms

@@

  Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service {[}TNRS; Boyle et al. (2013){]} provided by OpenTree (OpenTreeOfLife et al., n.d.). TNRS increases the probability of correctly finding the input taxon names in the chronogram database.

@@

At the time of writing of this manuscript (Mar 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN {[}@{]}. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.

@@

\includegraphics{../figures/figure1/figure1-horizontal.pdf}

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3673}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals (Kumar, Stecher, Suleski, \& Hedges, 2017).
Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O. Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education (Stoltzfus et al., 2013; Campbell O. Webb \& Donoghue, 2005).
We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user.
The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
Here we present the full implementation of DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/.
The latest implementation of the \texttt{datelife} R package is available on CRAN (v0.6.2; Sanchez-Reyes et al. (2022)) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize chronograms and compare them.

@@

  Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service (TNRS; Boyle et al. (2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.). TNRS increases the probability of correctly finding the input taxon names in the chronogram database.

@@

At the time of writing of this manuscript (Mar 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
Sanchez-Reyes, L. L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M. (2022). {datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}. \emph{R Package Version 0.6.2}. Retrieved from \url{https://doi.org/10.5281/zenodo.593938}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.

@@

\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
